---
title: "01a Introduction to R"
description: About R/RStudio installation, working with packages, info about current R sesssion
author:
   name: Marin Fotache
   url: https://github.com/marinfotache/Data-Processing-Analysis-Science-with-R
   affiliation: Alexandru Ioan Cuza University of Iasi, Dept. of Business Information Systms
   affiliation-url: https://www.feaa.uaic.ro/faculty/marin-fotache/
format:
   html:
      theme: flatly
      toc: true
      toc-depth: 4
      number-sections: true
editor: visual
execute: 
  echo: true
date: last-modified 
title-block-banner: '#053F56'
---

## Before you start

If you are an (ultra) beginner in R/RStudio, or if you managed to forget almost everything about R, here are some useful resources on basics (including installation on your laptop/desktop) of R and RStudio:

-   R Studio Education (Beginners) by Posit (former RStudio) <https://education.rstudio.com/learn/beginner/>

-   R resources (free courses, books, tutorials, & cheat sheets) by Paul van der Laken (a bit outdated, but still valuable) <https://paulvanderlaken.com/2017/08/10/r-resources-cheatsheets-tutorials-books/#introductory>

-   Our main reference during the sections related to basics of R, data import, data processing and data exploration will be **Hadley Wickham, Mine Çetinkaya-Rundel, Garrett Grolemund (2023). R for Data Science. O'Reilly** available at <https://r4ds.hadley.nz>.

-   To get an idea on what are we are trying to do in this section and subsequent ones, see <https://r4ds.hadley.nz/intro>.

## Install R and RStudio. Basic operations in RStudio

### Useful resources for R/RStudio installation and operations :

-   Chapter 1 Getting Started with Data in R. <https://moderndive.netlify.app/1-getting-started.html>

-   RStudio User Guide <https://docs.posit.co/ide/user/>.

### Install, load and use packages

Base R provides core features, but most of the R power and elegance resides in its packages (or libraries). A package must be installed just one (later, it can be updated, whenever necessary)

```{r}
#| label: install-lubridate
# package "lubridate" provides many elegant features for dealing with variables of type date, timestamps, intervals, etc.
# run only once, then comment it (with a hashtag)
#install.packages('lubridate', repos = "http://cran.us.r-project.org") 
```

Once a package is installed, it must be loaded into memory for every R session.

```{r}
#| label: load-lubridate

# now "install.packages" command is commented
# install.packages('lubridate') 

# before use it, be sure the package is loaded into memory 
# (next command will be run only once per R session)
library(lubridate) 

```

Next, we'll use a simple function provided by the ***lubridate*** package:

```{r}
#| label: use-lubridate

# Function "Sys.Data()" is provided by the base R (it displays the current date)
Sys.Date()

# we'll use function "wday" to display the current day of the week
lubridate::wday(Sys.Date(), label = TRUE) 

```

Note: Usually, it is not necessary to preceed the name of the function by its package, as in ***lubridate::wday***. Nevertheless, when two or more packages (loaded into the current sesssion memory) provide functions with identical names, in order to execute a specific version of the function, its name must be prefixed by its package name.

### Some packages are actually mega-packages

In all of the remaining sections, we'll use extensivery some packages included in the ***tidyverse*** ecosystem (***dplyr***, ***tidyr***, ***readr***, ***ggplot2***, etc.). Instead of installing them one by one, we will install all at once:

```{r}
#| label: install-and-load-tidyverse
#| echo: false

# next command will install all the packages included in the "tidyverse" (and will take a while)
#install.packages('tidyverse', repos = "http://cran.us.r-project.org") 

# "instal..." line must be commented after the installation is completed

# once per session, prior to the first call to a function in a "tidyverse" package, the ecosystem must be loaded into memory:
library(tidyverse)

```

### Install packages from other repositories

When not available on \[http://cran.us.r-project.org\] <http://cran.us.r-project.org>, some packages can be installed from other repositories (e.g., github); this is also the case when the development version is to be installed (instead the stable, "cran"-compliant one).

Next chunk will install the development version of ***ggstatsplot*** package:

```{r}
#install.packages('pak')
library(pak)
pak::pak("IndrajeetPatil/ggstatsplot")
library(ggstatsplot)
```

Next chunk will install the ***tidystringdist*** package from github:

```{r}
# install.packages(remotes)
remotes::install_github("ColinFay/tidystringdist")
```


### Avoiding package re-installation (when the package is already installed) 

```{r}

# specify package to be installed...
packages_to_be_installed <- c('glue', 'scales')

# install each of the two packages only if it is not yet installed
installed_packages <- packages_to_be_installed %in% rownames(installed.packages())
if (any(installed_packages == FALSE)) {
     install.packages(packages[!installed_packages])
}

# ... a simple solution would be
if (!require('glue')) install.packages('glue')



```


## Example of a chunk of R code which calls functions from multiple packages

Task: import the exchange rates published (daily) by the National Bank of Romania (BNR) and plot a basic chart with RON-EUR exchange rates for the most recent days:

```{r}
#| label: get-EUR-RON-exhange-rates
#| message: false

# load the required packages ("tidyverse is already loaded")
# library(tidyverse)
#install.packages(htmltab, repos = "http://cran.us.r-project.org")
library(htmltab)

# we get data from here
url <- 'http://www.bnr.ro/Exchange-rates-15192.aspx' 

# get the html table with the exchange rates
exchange_rates <- htmltab::htmltab(doc = url, 
     which = 1, encoding = "UTF-8")  

# take a look on the structure of the imported data
glimpse(exchange_rates) 

# look at the first five rows in the data
head(exchange_rates,5) 

```

Now, generate the simple plot in @fig-eur-ron :

```{r}
#| label: fig-eur-ron
#| fig-cap: RON-EUR exchange rates
ggplot2::ggplot(
     
     # first we specify the data source
     exchange_rates |> transmute(date = Data, EUR = as.numeric(EUR)), 
     
     #... then "aestetics", i.e. what to be displayed on the x and y axes
     aes(x = date, y = EUR)) +  
     
     # this id chart type (of the exchange rates)
     geom_point() + 
     
     # some features will increase lisibility
     theme(axis.text.x = element_text(angle = 45, hjust = 1)
     
     # end of the plot command            
     ) 

```

## Utilities: display all information about the current R session, installed packages, etc.:

### To display all information about the current R session:

```{r}
#| label: session-info
sessionInfo()

```

### To display all installed packages (not only those currently loaded into memory):

```{r}
#| code-fold: show

# with head(5) we'll limit the display to the first five paclages
installed.packages() |> head(5)
```

### Display directory where R is installed:

```{r}
.libPaths()
```

### Display the current (working) directory:

```{r}
getwd()
```

## How to continue from here? (Subsequent topics)

-   Variable data types (strings, numbers, dates, ...)

-   Data structures (vectors, data frames/tibbles, lists, ...)

-   Import data into R (from text/csv/spreadsheet files stored locally or on the internet; from databases; from other statistical packages)

-   Data processing with (mainly) tidyverse

-   Basic programming in R

-   Clean and process data, and deliver it for the analysis

-   Visualize data

-   Explore data (Exploratory Data Analysis) (requires also some statistical knowledge on statistical distributions, descriptive statistics, etc)

-   Infer.... (require also some statistical knowledge on statistical hypothesis testing)

-   Model data (require also some statistical knowledge on statistical modeling, such as linear regression, logistic regression, etc.)

-   Machine Learning (with the `tidymodels` ecosystem)

-   Work with Big Data tools

Other sections of interest:

-   XML and JSON file management

-   Web scraping
